---
title: "CUBClustR User Manual"
author: "Matteo Ventura"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CUBClustR User Manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

## Introduction
The MLC-CUB model extends the CUB (Combination of a Uniform and shifted Binomial) framework to multivariate rating data. It assumes that each response results from two latent components:

* **Feeling** – the respondent’s underlying attitude or preference;
* **Uncertainty** – the indecision inherent in the response process.

By combining these components within a model-based clustering approach estimated via the EM algorithm, MLC-CUB identifies groups of respondents sharing similar patterns of feeling and uncertainty.

## The 'univer' Dataset
The participants of this study were asked to evaluate five specific aspects of the University of Naples Federico II Orientation services. Each item corresponds to a key dimension of the student experience:

* **Q1 – Information:** Satisfaction regarding the quality and clarity of the acquired information.
* **Q2 – Willingness:** Satisfaction regarding the availability and helpfulness of the staff.
* **Q3 – Office Hours:** Satisfaction regarding the accessibility and convenience of the opening hours.
* **Q4 – Competence:** Satisfaction regarding the professional expertise and knowledge of the staff.
* **Q5 – Global Satisfaction:** An overall evaluation of the orientation service experience.

The responses were collected using a seven-point Likert scale (1 = "Very unsatisfied", 7 = "Extremely satisfied").

## 1. Data Preparation

First, let's load the required packages and the dataset to prepare the matrix of responses.

```{r setup}
library(CUBClustR)
library(CUB)
library(ggplot2)

data(univer)

# Select the 5 relevant ordinal items (columns 8 to 12)
R <- univer[, 8:12]
R <- apply(R, 2, as.integer)

# Maximum categories for each item (m = 7)
m_vector <- rep(7, ncol(R))
```

## 2. Model Selection (BIC)

We test from 1 to 7 latent classes to find the optimal balance between model complexity and fit. The `compare_MLCCUB_models` function automatically computes the Bayesian Information Criterion (BIC) for each model.

*(Note: Depending on your hardware, estimating multiple models might take a few moments).*

```{r model-selection, results='hide'}
# For the vignette, we use reduced parameters to speed up the compilation.
# In a real analysis, EM_iter = 10, max_iter = 1500, and a wider K_vector are recommended.
model_selection <- compare_MLCCUB_models(
  data     = R, 
  m        = m_vector, 
  K_vector = 1:5, 
  EM_iter  = 1, 
  max_iter = 50
)
```

We can plot the BIC trend to visualize the optimal number of clusters:

```{r bic-plot}
ggplot(model_selection$summary, aes(x = K, y = BIC)) +
  geom_line(color = "#2C3E50", linewidth = 1.2) +
  geom_point(color = "#BD4F6C", size = 3) +
  theme_bw() +
  labs(title = "Model Selection: BIC Trend (University Data)",
       x     = "Number of Latent Classes (K)",
       y     = "BIC") +
  scale_x_continuous(breaks = 1:5)
```

## 3. Stability & Identifiability Check

Assuming the BIC results suggest a 3-cluster solution ($K=3$), we evaluate its identifiability using a bootstrap procedure. The function returns the Adjusted Rand Index (ARI) distribution: values near 1 indicate highly stable clusters.

This chunk of code is not evaluated to speedup the building of the vignette.
```{r identifiability, eval = F}
# We use n_boot = 5 and n_cores = 1 to prevent timeouts and system permission blocks
identifiability <- check_identifiability(
  data      = R, 
  m         = m_vector, 
  K         = 3, 
  n_boot    = 100, 
  n_cores   = 4 
)

# Display the ARI distribution
identifiability$plot
```

## 4. Final Model Visualization (K = 3)

Let's extract the 3-cluster model from our model selection list to interpret the results.

```{r extract-model}
model_k3 <- model_selection$all_models$K3
```

### Parameter Space
The first visualization allows us to identify "satisfied", "unhappy", or "uncertain" groups by plotting the Feeling vs. Uncertainty parameters.

```{r plot-params}
plot_params <- plot_MLCCUB_clusters(model_k3)
plot_params + labs(title = "University Clusters: Parameter Space")
```

### Observed vs Theoretical Fit
We can also visually check how well the model mimics the real data distributions for each item and cluster.

```{r plot-fit, fig.width=8, fig.height=6}
plot_fit <- plot_CUB_fit(
  model_obj  = model_k3, 
  data       = R, 
  m          = m_vector,
  line_color = "#93B5C6"
)
plot_fit
```

## 5. Goodness of Fit Diagnostics

For a quantitative assessment of the fit quality, we compute the Weighted Average Dissimilarity (WAD). A global index closer to 0 indicates a perfect fit.

```{r wad-diagnostics}
fit_metrics <- calculate_WAD(model_obj = model_k3, data = R, m = m_vector)

# Diagnostic results
fit_metrics$WAD                   # Global index 
fit_metrics$Cluster_Dissimilarity # Unweighted average dissimilarity by cluster
fit_metrics$Item_Dissimilarity    # Weighted average dissimilarity by item
```

## 6. Non-metric Multidimensional Scaling (NMDS)

Finally, we use NMDS to project respondents into a 2D space. Unlike classical MDS, NMDS preserves the rank-order of distances, making it ideal for non-linear ordinal rating data. We use the Manhattan distance on a subsample of respondents for clarity.

```{r nmds, results='hide'}
library(vegan)

set.seed(123) 
idx_sample <- sample(1:nrow(R), 800) 
R_sub <- R[idx_sample, ]
cluster_sub <- model_k3$class[idx_sample]

nmds_res <- metaMDS(R_sub, 
                    distance = "manhattan", 
                    k = 2, 
                    trymax = 20,
                    autotransform = FALSE, 
                    trace = 0)
```

```{r nmds-plot}
cat("NMDS Stress Value:", round(nmds_res$stress, 4), "\n")

nmds_data <- data.frame(
  NMDS1   = nmds_res$points[, 1],
  NMDS2   = nmds_res$points[, 2],
  Cluster = factor(cluster_sub)
)

my_palette <- c("#93B5C6", "#DDEDAA", "#D7816A", "#F0CF65", "#BD4F6C")

ggplot(nmds_data, aes(x = NMDS1, y = NMDS2, color = Cluster)) +
  geom_jitter(width = 0.3, height = 0.3, alpha = 0.8, size = 2) +
  scale_color_manual(values = my_palette) +
  theme_minimal() +
  labs(
    title = "Non-metric MDS: Respondents clusters",
    x = "Dimension 1",
    y = "Dimension 2"
  ) +
  theme(plot.title = element_text(face = "bold"))
```

## Reference
Ventura M., Jacques J., Zuccolotto P. (2025), *Model-based Clustering of Multivariate Rating Data accounting for Feeling and Uncertainty*. Journal of Classification. 1-22.